Correct and Secure Web Programming using Dependent Types and Embedded Domain-Specific Languages

### 2013-07-30 paper/main.tex
More work on paper

\usepackage{amsmath}
\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usetikzlibrary{arrows,automata}
\WarningsOff

\begin{document}
\newcommand{\idris}{\textsc{Idris}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

@@

\section{Introduction}
Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits such as SQL injection or cross-site scripting. Many web applications are written in dynamically-checked scripting languages \cite{}, which allow rapid development. A significant drawback, however, is that such languages do not provide  the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.

%%------- EXAMPLE
Allow us to consider a simple, traditional database access routine, written in PHP, where we wish to obtain the name and address of every employee working in a given department, \texttt{\$dept}. We firstly construct an object representing a database connection, where the arguments are the database host, user, password and name respectively:

{\small
\begin{verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{verbatim}
}

We must subsequently check to see if the connection was successful, and cease execution of the database routine if so. Of course, this check is optional, so it would be perfectly possible to omit the check: this would, however, cause problems with later execution steps.

{\small
\begin{verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{verbatim}
}

We then create a prepared statement detailing our query, and bind the \texttt{`dept'} value:

{\small
\begin{verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param($dept);
\end{verbatim}
}

After the parameters have been bound, we must execute the statement, assign variables into which results will be stored, and fetch each row in turn. At this point, a developer may forget to execute the statement before beginning to fetch results or may have forgotten to bind the parameters to the statement:

{\small
\begin{verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);

  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{verbatim}
}

Finally, once the statement and connection are no longer needed, they should be closed in order to discard the associated resources:

{\small
\begin{verbatim}
  $stmt->close();
  $conn->close();
\end{verbatim}
}

Even in this small example, there exists a very specific resource usage protocol. Firstly, a connection to the database must be opened. The object-oriented style used in the example encapsulates this to an extent, as the object must be created in order for operations to be performed, however it is less obvious in a procedural version of the code. Secondly, a prepared statement is created, using the raw SQL and placeholders to which variables are later bound. The statement is then executed, and each row is retrieved from the database. Finally, the resources are freed.

@@

An effect is represented as an instance of the \texttt{EFFECT} type. Such instances may be associated with a \textit{resource}, 
Describing a program in the \texttt{Effects} library is achieved using the following data type:
{\small
\begin{verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) ->
       (es’ : List EFFECT) ->
       (a : Type) -> Type
\end{verbatim}
}
\texttt{EffM} is a type that is parameterised over an execution context (\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list of effects at the end of the program's execution (\texttt{es'}), and a return type (\texttt{a}). The list of input types can be used to ensure that an operation may only be called when the program is in a particular state, and since the list of input effects may differ from the list of output effects, we may modify the resource associated with the state after each operation. This allows us to simply enforce a resource usage protocol, and deviation from this protocol will result in a compile-time error.

Implementing a new effect involves creating a new algebraic data type, of kind \texttt{Effect}, which describes the available operations and how they modify the resource associated with the effect. Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
{\small
%TODO: Execution
\begin{verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{verbatim}
}
Operations are made available to users by writing functions with the corresponding \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
{\small
\begin{verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{verbatim}
}
Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context. Finally, \texttt{Effect} instances are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff} constructor, which is provided with the initial resource value.

@@

By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI process as the unit type, ().
{\small
\begin{verbatim}
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{verbatim}
}
Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses. Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.

@@

{\small
\begin{verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{verbatim}
}
Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.

@@

In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We may therefore define the following type synonym, which we may use in order to make use of the CGI framework: 
{\small
\begin{verbatim}
CGIProg : List EFFECT -> Type -> Type
CGIProg effs a = 
 Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
\end{verbatim}
}
This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.

@@

{\small
\begin{verbatim}
sayHello : CGIProg [] ()
sayHello = output "Hello, world!"

main : IO ()
main = do runAction [initCGIState] sayHello
          pure ()
\end{verbatim}
}
% =================================================

@@

Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library. There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query. This is reflected in the associated resource, which again is parameterised by the current protocol state.
{\small
\begin{verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

@@

data StmtPointer = ValidStmt Ptr
                 | InvalidStmt 
\end{verbatim}
}
If a failure happens at any point during the computation, the resource is updated to reflect the failure. At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety. The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.

%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}. In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function. Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function. We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor. 
{\small
\begin{verbatim}
data StepResult = Unstarted
                | StepFail
                | StepComplete
                | NoMoreRows
\end{verbatim}
}
Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource. Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore statically guarantee that only calls that will return a valid result are executed.

@@

We define a function of type:
{\small
\begin{verbatim}
String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))
\end{verbatim}
}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned. 
{\small
\begin{verbatim}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err $
\end{verbatim}
}
The program initially attempts to open a connection to the \texttt{people.db} database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.

@@

\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database. This function is then recursively called until there are no more rows to process.
{\small
\begin{verbatim}
collectResults : 
  Eff IO 
  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
  (List (String, Int))
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return [] $
\end{verbatim}
}

In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
{\small
\begin{verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{verbatim}
}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
{\small
\begin{verbatim}
  ResultSet : Type
  ResultSet = List (List DBVal)
\end{verbatim}
}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
{\small
\begin{verbatim}
executeSelect : String ->
                String -> 
                List (Int, DBVal) -> 
                (Eff m [SQLITE 
                  (SQLiteRes PreparedStatementExecuting)
                   ] 
                 (List DBVal)
                ) -> 
                Eff m [SQLITE ()] (Either String ResultSet)
\end{verbatim}
}
% =================================================

\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

@@

We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session. A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.

%TODO: Fix up the labels
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]

@@

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
\end{tikzpicture}
}
\caption{Session Handler Resource Usage Protocol}
\label{fig:sessionstates}
\end{figure}

Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler. In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key. The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage. This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.

Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
{\small
\begin{verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

data SessionRes : SessionStep -> Type where
  InvalidSession : SessionRes s  
  ValidSession   : SessionID -> 
                   SessionData -> 
                   SessionRes s
\end{verbatim}
}
The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.

@@

{\small
\begin{verbatim}
SELECT `name`, `address` FROM `staff` LIMIT ' + page + ', 5
\end{verbatim}
}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

@@

{\small
\begin{verbatim}
using (G : List FormTy, E : List WebEffect)
  data FormRes : List FormTy -> 
                 List WebEffect -> Type where
    FR : Nat -> 
         List FormTy -> 
         List WebEffect -> 
         String -> 
         FormRes G E

@@

\end{verbatim}
}

@@

{\small
\begin{verbatim}
handleNewThread : Maybe String -> 
                  Maybe String -> 

>>>>

\usepackage{amsmath}
\usepackage{silence}
\usepackage{pgf}
\usepackage{tikz}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usetikzlibrary{arrows,automata}
\WarningsOff

\begin{document}
\fvset{fontsize=\small}
\newcommand{\idris}{\textsc{Idris}}
\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

@@

\section{Introduction}
Web applications, whilst ubiquitous, are also prone to incorrect construction and security exploits such as SQL injection \cite{owasp:sqli} or cross-site scripting \cite{owasp:xss}. Security breaches using such exploits are far-reaching, and high profile cases involve large corporations such as Sony, who suffered a well-publicised and extremely costly SQL Injection breach in 2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012 \cite{imperva:yahoo}.

Many web applications are written in dynamically-checked scripting languages such as PHP, Ruby or Python, which facilitate rapid development \cite{w3techs:webpls}. A significant drawback, however, is that such languages do not provide  the same static guarantees about runtime behaviour afforded by programs with more expressive, static type systems, instead relying on extensive unit testing to ensure correctness and security.

%%------- EXAMPLE
Allow us to consider a simple, traditional database access routine, written in PHP, where we wish to obtain the name and address of every employee working in a given department, \texttt{\$dept}. We firstly construct an object representing a database connection, where the arguments are the database host, user, password and name respectively:
%
\begin{Verbatim}
$conn = new mysqli("localhost", "username", 
                    "password", "db");
\end{Verbatim}
%
We must subsequently check to see if the connection was successful, and cease execution of the database routine if so. Of course, this check is optional, so it would be perfectly possible to omit the check: this would, however, cause problems with later execution steps.
%
\begin{Verbatim}
  if (mysqli_connect_errno()) {
      exit();
  }
\end{Verbatim}
%
We then create a prepared statement detailing our query, and bind the \texttt{`dept'} value:
\begin{Verbatim}
  $stmt = $conn->prepare("SELECT `name`, `address` 
     FROM `staff` WHERE `dept` = ?);
  $stmt->bind_param('s', $dept);
\end{Verbatim}
%
After the parameters have been bound, we must execute the statement, assign variables into which results will be stored, and fetch each row in turn. At this point, a developer may forget to execute the statement before beginning to fetch results or may have forgotten to bind the parameters to the statement:
%
\begin{Verbatim}
  $stmt->execute();
  $stmt->bind_result($name, $address);
  while ($stmt->fetch()) {
    printf("Name: %s, Age: %s", $name, $age);
  }
\end{Verbatim}
%
Finally, once the statement and connection are no longer needed, they should be closed in order to discard the associated resources:
%
\begin{Verbatim}
  $stmt->close();
  $conn->close();
\end{Verbatim}
%
Even in this small example, there exists a specific resource usage protocol. Firstly, a connection to the database must be opened. The object-oriented style used in the example encapsulates this to an extent, as the object must be created in order for operations to be performed, however it is less obvious in a procedural version of the code. Secondly, a prepared statement is created, using the raw SQL and placeholders to which variables are later bound. The statement is then executed, and each row is retrieved from the database. Finally, the resources are freed.

@@

An effect is represented as an instance of the \texttt{EFFECT} type. Such instances may be associated with a \textit{resource}, 
Describing a program in the \texttt{Effects} library is achieved using the following data type:
\begin{Verbatim}
EffM : (m : Type -> Type) ->
       (es : List EFFECT) ->
       (es’ : List EFFECT) ->
       (a : Type) -> Type
\end{Verbatim}
\texttt{EffM} is a type that is parameterised over an execution context (\texttt{m}), a list of effects that the program may use (\texttt{xs}), a list of effects at the end of the program's execution (\texttt{es'}), and a return type (\texttt{a}). The list of input types can be used to ensure that an operation may only be called when the program is in a particular state, and since the list of input effects may differ from the list of output effects, we may modify the resource associated with the state after each operation. This allows us to simply enforce a resource usage protocol, and deviation from this protocol will result in a compile-time error.

Implementing a new effect involves creating a new algebraic data type, of kind \texttt{Effect}, which describes the available operations and how they modify the resource associated with the effect. Each operation consists of an initial resource \texttt{res}, the resource at the end of the operation \texttt{res'}, and the output type of the operation \texttt{t}.
%TODO: Execution
\begin{Verbatim}
Effect : Type
Effect = (res : Type) -> (res’ : Type) ->
         (t : Type) -> Type
\end{Verbatim}
Operations are made available to users by writing functions with the corresponding \texttt{EffM} type, which may then be used to invoke the operation from within an effectful program.
\begin{Verbatim}
queryPostVar : 
 String -> 
 Eff m [CGI (InitialisedCGI TaskRunning)] 
   (Maybe String)
queryPostVar query = (QueryPostVar query)
\end{Verbatim}
Additionally, we define \textit{handlers}, which describe how each operation is interpreted in a given execution context. Finally, \texttt{Effect} instances are promoted to \texttt{EFFECT} instances through the use of the \texttt{MkEff} constructor, which is provided with the initial resource value.

@@

By modelling CGI as a resource-dependent algebraic effect, we may enforce a resource usage protocol which, even though the program may be running in an IO execution context, prevents arbitrary IO from being performed and therefore ensures that the headers are written correctly. In order to accomplish this, we define an effect, \texttt{Cgi}, and an associated resource, \texttt{InitialisedCGI}, which is parameterised over the current state, \texttt{CGIStep}. This resource describes the current state, alongside a \texttt{CGIInfo} record which contains information from the request. We represent an uninitialised CGI process as the unit type, ().
\begin{Verbatim}[samepage]
data CGIStep = Initialised 
             | TaskRunning 
             | TaskCompleted 
             | HeadersWritten 
             | ContentWritten

data InitialisedCGI : CGIStep -> Type where
  ICgi : CGIInfo -> InitialisedCGI s
\end{Verbatim}
Figure ~\ref{fig:cgistates} shows the states through which the CGI program progresses. Each operation performed in an effectful program requires the resource to be of a certain type, and the completion of the operation may alter the type or value of the resource.

@@

\begin{Verbatim}
data Cgi : Effect where
    Init : Cgi () (InitialisedCGI Initialised) ()
    StartRun : Cgi (InitialisedCGI Initialised) 
                   (InitialisedCGI TaskRunning) ()
    FinishRun : Cgi (InitialisedCGI TaskRunning) 
                    (InitialisedCGI TaskCompleted) ()
    WriteHeaders : Cgi (InitialisedCGI TaskCompleted) 
                       (InitialisedCGI HeadersWritten) ()
    WriteContent : Cgi (InitialisedCGI HeadersWritten) 
                       (InitialisedCGI ContentWritten) ()
    OutputData : String -> 
                 Cgi (InitialisedCGI TaskRunning) 
                     (InitialisedCGI TaskRunning) ()
    RunAction : Env IO (CGI (InitialisedCGI TaskRunning) 
                               :: effs) -> 
                       CGIProg effs a -> 
                       Cgi (InitialisedCGI TaskRunning) 
                           (InitialisedCGI TaskRunning) a
\end{Verbatim}
Additional operations, including those to query POST and GET variables, are omitted in the interest of brevity.

@@

In \idris{}, types are first-class, meaning that they may be treated like other terms in computations. We may therefore define the following type synonym, which we may use in order to make use of the CGI framework: 
\begin{Verbatim}
CGIProg : List EFFECT -> Type -> Type
CGIProg effs a = 
 Eff IO (CGI (InitialisedCGI TaskRunning) :: effs) a
\end{Verbatim}
This is then passed, along with initial values for other effects that the user may wish to use, to the runAction function, which invokes the RunAction operation and executes the user-specified action.

@@

\begin{Verbatim}
import Cgi
import Effects

sayHello : CGIProg [] ()
sayHello = output "Hello, world!"

main : IO ()
main = do runAction [initCGIState] sayHello
          pure ()
\end{Verbatim}
% =================================================

@@

Figure ~\ref{fig:sqlitestates} shows a resource usage protocol for database access, which we have implemented for the SQLite library. This is encapsulated by the \texttt{Sqlite} effect; we again omit some operations, such as those to bind and retrieve data types other than \texttt{String}, in the interest of brevity.

\begin{Verbatim}[samepage]
data Sqlite : Effect where
  OpenDB : 
    String -> 
    Sqlite () 
           (SQLiteRes ConnectionOpened) Bool
  CloseDB : 
    Sqlite (SQLiteRes ConnectionOpened) () Bool
  PrepareStatement : 
    String -> 
    Sqlite (SQLiteRes ConnectionOpened) 
           (SQLiteRes PreparedStatementOpen) Bool
  StartBind : 
    Sqlite (SQLiteRes PreparedStatementOpen) 
           (SQLiteRes PreparedStatementBinding) ()
  BindText : 
    ArgPos -> String -> Int -> 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBinding) 
           Bool
  FinishBind : 
    Sqlite (SQLiteRes PreparedStatementBinding) 
           (SQLiteRes PreparedStatementBound) 
           Bool
  ExecuteStatement : 
    Sqlite (SQLiteRes PreparedStatementBound) 
           (SQLiteRes PreparedStatementExecuting) ()
  Finalise : 
    Sqlite (SQLiteRes PreparedStatementExecuting) 
           (SQLiteRes ConnectionOpened) 
           Bool
  GetColumnText : 
    Int -> 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           String
  RowStep : 
    Sqlite (SQLiteRes PreparedStatementExecuting)
           (SQLiteRes PreparedStatementExecuting)
           StepResult
\end{Verbatim}
There are three main phases involved in the usage of the SQLite protocol: connection to the database, preparation of a query, and execution of the query. This is reflected in the associated resource \texttt{SQLiteRes}, which again is parameterised by the current protocol state.
\begin{Verbatim}
data SQLiteRes : Step -> Type where
  OpenConn : DBPointer -> SQLiteRes s
  OpenStmt : DBPointer -> StmtPointer -> SQLiteRes s
  ExecutingStmt : DBPointer -> 
                  StmtPointer -> 
                  StepResult -> 
                  SQLiteRes s

@@

data StmtPointer = ValidStmt Ptr
                 | InvalidStmt 
\end{Verbatim}
If a failure happens at any point during the computation, the resource is updated to reflect the failure: if, for example, the library failed to create a connection to the database, the resource value would be updated to \texttt{OpenConn InvalidConn}. At this point, no further side-effecting requests are made to the underlying SQLite library, in order to ensure safety. The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and \texttt{executeFail} utility functions allow for failures, once detected, to be handled by executing the appropriate sequence of state transition functions to dispose of any open resources and return to the initial protocol state.

%Queries are evaluated through one or more calls to the \texttt{nextRow} function, which either executes an update statement or returns the next row of a result set. 
SQL queries are evaluated in SQLite upon a call to the C library function \texttt{sqlite3\_step()}. In the case that a statement returns a result set, each subsequent call retrieves another row for processing using a column access function. Once all rows have been retrieved, the library returns \texttt{SQLITE\_DONE}, meaning that no further calls should be made without resetting the function. We encapsulate this requirement through the \texttt{StepResult} data type within the \texttt{ExecutingStmt} constructor. 
\begin{Verbatim}
data StepResult = Unstarted
                | StepFail
                | StepComplete
                | NoMoreRows
\end{Verbatim}
Each call to \texttt{nextRow}, which is a wrapper around the underlying \texttt{sqlite3\_step()} C library function, returns a result of type \texttt{StepResult}, which is then reflected in the resource. Calls to \texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult} is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore statically guarantee that only calls that will return a valid result are executed.

@@

We define a function of type:
\begin{Verbatim}
String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))
\end{Verbatim}
This means that the program will be run in the IO execution context, and must start and end with no active resources. The return type indicates that either a list of \text{(String, String)} pairs, representing names and addresses in the database, or an error will be returned. 
\begin{Verbatim}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, Int)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err $
\end{Verbatim}
The program initially attempts to open a connection to the \texttt{people.db} database. At this point, since the \texttt{OpenDB} operation has been invoked, the program transitions to the \texttt{ConnectionOpened} state. The \texttt{openDB} function returns a Boolean value indicating whether or not the operation is successful. If not, then the \texttt{connFail} function is called to generate an appropriate error and dispose of the resources, as shown in ~\ref{fig:sqlitestates}.

@@

\texttt{CollectResults} is a simple function which makes a call to \texttt{nextRow} in order to make the next row of the result set available for processing, and uses the \texttt{getColumnText} and \texttt{getColumnInt} functions to retrieve the data from the database. This function is then recursively called until there are no more rows to process.
\begin{Verbatim}
collectResults : 
  Eff IO 
  [SQLITE (SQLiteRes PreparedStatementExecuting)] 
  (List (String, Int))
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return [] $
\end{Verbatim}
In order to decrease unnecessary boilerplate code in user applications, we provide functions which abstract out unnecessary parts of this process. In order to do this, we define the algebraic data type \texttt{DBVal}, which is a tagged union over simple primitive types:
\begin{Verbatim}
data DBVal = DBInt Int
           | DBText String
           | DBFloat Float
           | DBNull
\end{Verbatim}
We also make use of the \texttt{ResultSet} type, which is a list of returned database rows.
\begin{Verbatim}
  ResultSet : Type
  ResultSet = List (List DBVal)
\end{Verbatim}
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given a query, a list of variables to bind and their associated indices within the query, and a function which is used to extract information out of each returned database row, returns either a \texttt{ResultSet} or an error.
\begin{Verbatim}
executeSelect : 
  String -> String -> List (Int, DBVal) -> 
  (Eff m [SQLITE (SQLiteRes PreparedStatementExecuting)] 
                 (List DBVal)) -> 
  Eff m [SQLITE ()] (Either String ResultSet)
\end{Verbatim}
% =================================================
\subsection{A Simple Session Handler}
More complex web applications require some persistent state across separate requests. This is often done through an abstraction of a \textit{session}, wherein a cookie is set on the remote host containing a unique session ID, which is in turn used to retrieve data. In this section, we describe the implementation of a simple session handler, and the resource usage protocol involved.

@@

We implement this by having a SQLite database, containing two tables: \texttt{session}, which stores session keys and their associated expiry dates, and \texttt{sessiondata}, which contains the data associated with each session. A datum associated with the session is described as a tagged union containing one of the primitive types \texttt{String}, \texttt{Bool} or \texttt{Int}, which is serialised alongside a type tag for storage in the database.
%TODO: Fix up the labels
\begin{Verbatim}[samepage]
data Session : Effect where
  LoadSession : 
  SessionID -> 
  Session (SessionRes SessionUninitialised) 
          (SessionRes SessionInitialised) 
          (Maybe SessionData)
  UpdateSession : 
    SessionData -> 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) ()
  CreateSession : 
    SessionData -> 
    Session (SessionRes SessionUninitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  DeleteSession : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool 
  WriteToDB : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) Bool
  DiscardSessionChanges : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionUninitialised) ()
  GetSessionID : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionID)
  GetSessionData : 
    Session (SessionRes SessionInitialised) 
            (SessionRes SessionInitialised) 
            (Maybe SessionData)
\end{Verbatim}
\begin{figure}[htpb!]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=7.5em, font={\sffamily\bfseries}]

@@

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=-1cm, text width=3cm] {\texttt{loadSession, createSession}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, xshift=6em, text width=3cm] {\texttt{writeToDB, discardSessionChanges, deleteSession}} (A)
            edge[stateEdge, loop left]   node[edgeLabel] {\texttt{updateSession}} (B);
\end{tikzpicture}
}
\caption{Session Handler Resource Usage Protocol}
\label{fig:sessionstates}
\end{figure}
Figure ~\ref{fig:sessionstates} shows the resource usage protocol associated with the session handler. In this application, there exist two states: \texttt{SessionUninitialised}, in which the user may load or create a new session, and \texttt{SessionInitialised}, in which the user may update the representation of the session in memory, serialise the session and write it to the database, or delete the session and invalidate the user's session key. The introduction of these two states ensures that changes are explicitly either written or discarded, eliminating the possibility of a developer updating the session but neglecting to commit it to persistent storage. This, of course, is under the assumption that the process exits cleanly: we attempt to facilitate this by writing total functions where possible.

Much like the SQLite effect, we encapsulate failure by reflecting it in the resource associated with the effect. 
\begin{Verbatim}
data SessionStep = SessionUninitialised
                 | SessionInitialised

data SessionRes : SessionStep -> Type where
  InvalidSession : SessionRes s  
  ValidSession   : SessionID -> 
                   SessionData -> 
                   SessionRes s
\end{Verbatim}
The \texttt{SessionRes} data type is parameterised over the current state, which determines which operations may be performed, and has two constructors: \texttt{InvalidSession} and \texttt{ValidSession}. If an operation such as creating a new session fails, no further side-effecting calls will be made, in turn preserving integrity.

@@

\begin{Verbatim}
'SELECT `name`, `address` FROM `staff` LIMIT ' + page + '
                                                     , 5'
\end{Verbatim}
The \texttt{page} variable is assumed to be an integer, but may instead be modified by an attacker to include a malicious string which would alter the semantics of the query, allowing an attacker to execute a blind SQL injection attack. % Might be a good idea to cite an SQL injection paper which uses LIMIT clauses here

@@

\begin{Verbatim}
using (G : List FormTy, E : List WebEffect)
  data FormRes : List FormTy -> 
                 List WebEffect -> Type where
    FR : Nat -> 
         List FormTy -> 
         List WebEffect -> 
         String -> 
         FormRes G E

@@

\end{Verbatim}

@@

\begin{Verbatim}
handleNewThread : 
  Maybe String -> Maybe String -> 
  FormHandler [CGI (InitialisedCGI TaskRunning), 
